Transmembrane helices are an overlooked and evolutionarily conserved source of major histocompatibility complex class I and II epitopes

### 2021-12-12 frontiers.tex
'frontiers.pdf' is done

We define a protein to be a binder if, for a certain MHC allele, 
any of its 9-mer or 14-mer peptides have an IC50 value in the lowest 2\% of 
all peptides within a \emph{proteome} 
(see supplementary Tables \ref{tab:ic50_binders_mhc1} and \ref{tab:ic50_binders_mhc2} for values), 
this differs from our previous study where we defined
a binder as having an IC50 in the lowest 2\% 
of the peptides within a \emph{protein}.
% See https://github.com/richelbilderbeek/bianchi_et_al_2017/blob/72e6755a31d400158368509fd80a41e984677ab1/predict-binders.R#L17
This revised definition precludes bias of proteins 
that give rise to no or only very few MHC epitopes.
To verify that the slight change in method yields similar results,
a side by side comparison is shown in the supplementary materials, 
Figures \ref{fig:bianch_et_al_2017_1a} and \ref{fig:bilderbeek_et_al_2021_1a}.

@@

The number of SNPs was limited to the first 250 variations per gene,
resulting in $\approx$61k variations.
Only variations that result in a SNP for
a single amino acid substitution were analyzed, resulting in $\approx$38k SNPs.
The exact amounts can be found in the supplementary materials,
Tables \ref{tab:ncbi_counts_1} and \ref{tab:ncbi_counts_2}.

% \paragraph{Selection of SNPs}
%
SNPs were picked based on ID number, which is linked to their discovery date. To verify that these ID numbers are unrelated to SNP positions, the relative positions of all analyzed SNPs in a protein were determined. This analysis showed no positional bias of the SNPs, as shown in supplementary figure \ref{fig:snp_rel_pos}.

@@

We next wondered if the over-representation of TMH-derived peptides 
would also be confirmed for MHC-II. 
Figure \ref{fig:1}A shows the percentages of MHC-II epitopes 
predicted to be overlapping with TMHs for our human, viral and bacterial proteomes.
We found that TMH-derived peptides are over-presented in all
of the 21 MHC-II alleles, 
for a human, bacterial and viral proteome,
except for \verb;HLA-DRB3*0101; in \emph{M. tuberculosis}.
See supplementary Table \ref{tab:tmh_binders_mhc2} 
for the exact TMH and epitope counts.

@@

Figure (\ref{fig:2a} and \ref{fig:thm_presentation_correlation}) shows there are similar levels of
over-presentation of TMH-derived epitopes between (1) the
percentage of TMH-derived epitopes that is reported in the IEDB database
versus (2) the percentage of TMH-derived epitopes that is predicted to be presented
in MHC-I alleles.
For MHC-II alleles, there were too few epitopes per MHC allele
to result in an informative figure.

In figure \ref{fig:2b} we grouped all the epitopes 
presented by MHC-I and MHC-II alleles 
by the percentage of TMH-derived epitopes,
which are 22\% and 10\%, respectively.

These findings robustly confirm that
epitopes derived from human TMHs are presented in both MHC-I and MHC-II, and support that they are over-presented.
See the supplementary Table \ref{tab:elution} for the exact values.

We also mined the IEDB database for epitopes for
any type of T-cell response from the specified
alleles, from the total reports 36\% and 7\% 
concerned TMH-derived epitopes in MHC class I and II, respectively 
(see Figure \ref{fig:t_cells_present_tmh_derived_epitopes}).

@@

We addressed the question whether there is an evolutionary advantage in presenting TMHs.
We determined the conservation of TMHs 
by comparing the occurrences of SNPs located in TMHs or soluble protein regions 
for the genes coding for membrane proteins.
We obtained 911 unique gene names associated with the phrase 'membrane protein',
which are genes coding for both membrane-associated proteins (MAPs, which have no TMH) and 
transmembrane proteins (TMPs, which have at least one TMH).
These genes are linked to 4,780 protein isoforms, 
of which 2,553 are predicted to be TMPs and 
2,237 proteins are predicted to be MAPs.
We obtained 37,630 unique variations, 
of which 9,621 are SNPs that resulted in a straightforward amino acids substitution, 
of which 6,062 were located in predicted TMPs.
See supplementary Tables \ref{tab:ncbi_counts_1} and \ref{tab:ncbi_counts_2} 
for the detailed numbers and distributions of SNPs.

@@

%
% Figure 1
%
\begin{figure}[h!]
\begin{center}
\includegraphics[width=0.8\textwidth]{bbbq_article_figures_Page_1}% This is a *.eps file
\end{center}
\caption{
    \textbf{Over-presentation of TMH-derived epitopes on most MHC-I and -II alleles}
    \textbf{(A)} 
    The percentage of epitopes for MHC-I and -II alleles that are predicted to 
    overlap with TMHs for the proteomes of SARS-CoV-2 (top row), human (middle 
    row) and \emph{M. tuberculosis} (MtB; bottom row).
    The pair of horizontal red lines in each plot indicate the lower and upper bound 
    of the 99\% confidence interval.
    See supplementary Tables \ref{tab:tmh_binders_mhc2} and \ref{tab:tmh_binders_mhc1}
    for the exact TMH and  epitope counts.
    \textbf{(B-C)}
    Correlation between the percentages of predicted TMH-derived epitopes
    and the hydrophobicity score of all predicted epitopes for 
    human MHC-I \textbf{(B)} and MHC-II alleles \textbf{(C)}.
    Diagonal red line: linear regression analysis. 
    Labels are shorthand for the HLA alleles,
    see the supplementary Table \ref{tab:haplotype_abbreviations} for the names.}
  \label{fig:1}
\end{figure}
% 1A: \label{fig:bbbq_1_smart_results}
% 1B: \label{fig:hydrophobicity_1}
% 1C: \label{fig:hydrophobicity_2}

%
% Figure 2
%
\begin{figure}[h!]
\begin{center}
\includegraphics[width=\textwidth]{bbbq_article_figures_Page_2}% This is a *.eps file
\end{center}
\caption{
    \textbf{
      Analysis of epitope database shows that TMH derived epitopes are over presented.
    }
    The percentage of epitopes for MHC-I and -II alleles that overlap with TMHs
    that are presented. The pair of horizontal red lines in each plot indicate the
    lower and upper bound of the 99\% confidence interval. Note that only one line
    is visible as this interval is relatively narrow. 
    Alleles are listed in Table \ref{tab:haplotype_abbreviations}).
    \textbf{(A)} 
    Observed and predicted percentage of TMH-derived epitopes for MHC-I alleles. 
    \textbf{(B)} 
    MHC ligands from IEDB corresponding to TMH-derived epitopes. 
    The numbers above the bars denotes the number of TMH derived epitopes obtained.
}
\label{fig:elution}
\end{figure}
% Figure 2a: \label{fig:2a}
% Figure 2b: \label{fig:2b}

>>>>

We define a protein to be a binder if, for a certain MHC allele, 
any of its 9-mer or 14-mer peptides have an IC50 value in the lowest 2\% of 
all peptides within a \emph{proteome} 
(see supplementary Tables ST1 and ST2 for values), 
this differs from our previous study where we defined
a binder as having an IC50 in the lowest 2\% 
of the peptides within a \emph{protein}.
% See https://github.com/richelbilderbeek/bianchi_et_al_2017/blob/72e6755a31d400158368509fd80a41e984677ab1/predict-binders.R#L17
This revised definition precludes bias of proteins 
that give rise to no or only very few MHC epitopes.
To verify that the slight change in method yields similar results,
a side by side comparison is shown in the supplementary materials, 
Figures S1 and S2.

@@

The number of SNPs was limited to the first 250 variations per gene,
resulting in $\approx$61k variations.
Only variations that result in a SNP for
a single amino acid substitution were analyzed, resulting in $\approx$38k SNPs.
The exact amounts can be found in the supplementary materials,
Tables ST3 and ST4.

% \paragraph{Selection of SNPs}
%
SNPs were picked based on ID number, which is linked to their discovery date. To verify that these ID numbers are unrelated to SNP positions, the relative positions of all analyzed SNPs in a protein were determined. This analysis showed no positional bias of the SNPs, as shown in supplementary figure S3.

@@

We next wondered if the over-representation of TMH-derived peptides 
would also be confirmed for MHC-II. 
Figure \ref{fig:1}A shows the percentages of MHC-II epitopes 
predicted to be overlapping with TMHs for our human, viral and bacterial proteomes.
We found that TMH-derived peptides are over-presented in all
of the 21 MHC-II alleles, 
for a human, bacterial and viral proteome,
except for \verb;HLA-DRB3*0101; in \emph{M. tuberculosis}.
See supplementary Table ST5 
for the exact TMH and epitope counts.

@@

Figure (\ref{fig:2}A and S4) shows there are similar levels of
over-presentation of TMH-derived epitopes between (1) the
percentage of TMH-derived epitopes that is reported in the IEDB database
versus (2) the percentage of TMH-derived epitopes that is predicted to be presented
in MHC-I alleles.
For MHC-II alleles, there were too few epitopes per MHC allele
to result in an informative figure.

In figure \ref{fig:2}B we grouped all the epitopes 
presented by MHC-I and MHC-II alleles 
by the percentage of TMH-derived epitopes,
which are 22\% and 10\%, respectively.

These findings robustly confirm that
epitopes derived from human TMHs are presented in both MHC-I and MHC-II, and support that they are over-presented.
See the supplementary Table ST6 for the exact values.

We also mined the IEDB database for epitopes for
any type of T-cell response from the specified
alleles, from the total reports 36\% and 7\% 
concerned TMH-derived epitopes in MHC class I and II, respectively 
(see Figure S5).

@@

We addressed the question whether there is an evolutionary advantage in presenting TMHs.
We determined the conservation of TMHs 
by comparing the occurrences of SNPs located in TMHs or soluble protein regions 
for the genes coding for membrane proteins.
We obtained 911 unique gene names associated with the phrase 'membrane protein',
which are genes coding for both membrane-associated proteins (MAPs, which have no TMH) and 
transmembrane proteins (TMPs, which have at least one TMH).
These genes are linked to 4,780 protein isoforms, 
of which 2,553 are predicted to be TMPs and 
2,237 proteins are predicted to be MAPs.
We obtained 37,630 unique variations, 
of which 9,621 are SNPs that resulted in a straightforward amino acids substitution, 
of which 6,062 were located in predicted TMPs.
See supplementary Tables ST3 and ST4 
for the detailed numbers and distributions of SNPs.

@@

%
% Figure 1
%
\begin{figure}[h!]
\begin{center}
\includegraphics[width=0.8\textwidth]{bbbq_article_figures_Page_1}% This is a *.eps file
\end{center}
\caption{
    \textbf{Over-presentation of TMH-derived epitopes on most MHC-I and -II alleles}
    \textbf{(A)} 
    The percentage of epitopes for MHC-I and -II alleles that are predicted to 
    overlap with TMHs for the proteomes of SARS-CoV-2 (top row), human (middle 
    row) and \emph{M. tuberculosis} (MtB; bottom row).
    The pair of horizontal red lines in each plot indicate the lower and upper bound 
    of the 99\% confidence interval.
    See supplementary Tables ST5 and ST7
    for the exact TMH and  epitope counts.
    \textbf{(B-C)}
    Correlation between the percentages of predicted TMH-derived epitopes
    and the hydrophobicity score of all predicted epitopes for 
    human MHC-I \textbf{(B)} and MHC-II alleles \textbf{(C)}.
    Diagonal red line: linear regression analysis. 
    Labels are shorthand for the HLA alleles,
    see the supplementary Table ST8 for the names.}
  \label{fig:1}
\end{figure}
% 1A: \label{fig:bbbq_1_smart_results}
% 1B: \label{fig:hydrophobicity_1}
% 1C: \label{fig:hydrophobicity_2}

%
% Figure 2
%
\begin{figure}[h!]
\begin{center}
\includegraphics[width=\textwidth]{bbbq_article_figures_Page_2}% This is a *.eps file
\end{center}
\caption{
    \textbf{
      Analysis of epitope database shows that TMH derived epitopes are over presented.
    }
    The percentage of epitopes for MHC-I and -II alleles that overlap with TMHs
    that are presented. The pair of horizontal red lines in each plot indicate the
    lower and upper bound of the 99\% confidence interval. Note that only one line
    is visible as this interval is relatively narrow. 
    Alleles are listed in Table ST8).
    \textbf{(A)} 
    Observed and predicted percentage of TMH-derived epitopes for MHC-I alleles. 
    \textbf{(B)} 
    MHC ligands from IEDB corresponding to TMH-derived epitopes. 
    The numbers above the bars denotes the number of TMH derived epitopes obtained.
}
\label{fig:2}
\end{figure}
% Figure 2a: \label{fig:2a}
% Figure 2b: \label{fig:2b}
